ODGI: scalable tools for pangenome graphs

### 2021-06-02 manuscript/figure_p.tex
caption: remove line on topology because you can not see it really
figure
Shortened
Fix table
edits

\begin{figure*}[ht]
    \vspace{-10pt}
    \begin{minipage}[b]{\textwidth}
        % \begin{center}

        \subfigure[aligned MHC pangenome]{
            \includegraphics[width=.48\textwidth]{fig/mhc-align.eps}
            \label{fig:1a}
        }
        \vspace{-11pt}
        % \newline
        \subfigure[consensus MHC pangenome]{
            \includegraphics[width=.48\textwidth]{fig/mhc-pangenome.eps}
            \label{fig:1b}
        }
        \caption{Pangenome visualisations of the Major Histocompatibility Complex (MHC) \textit{locus} by \odgi.
            (a) projection of full MHC \textit{locus} of haploid phased human genome
          assemblies, plus the chm13 cell line and GRCh38 reference genome (note the missing patch in the reference
          genome) as supplied by the
            Human Reference Pangenome Consortium\citep{HRPC} and created by \cmd{odgi build+sort+viz}. The coloured
            bars represent the linearised paths --- representing contigs as a zoomed out multi-sequence
            alignment. The black lines at the bottom show the graph topology.
            (b) consensus graph representation by \cmd{odgi build} of the same assemblies showing variations larger than $100$ base pairs.
            Loops or `bubbles' display contigs uniquely diverging from the consensus, caused by variation in repeats.
            The gene labels are super imposed by \cmd{odgi position} from a GRCh38 BED file and visualised by the Bandage tool\citep{26099265}.
        }
        \label{fig:1}
        % \end{center}
    \end{minipage}
\end{figure*}

>>>>

\begin{figure*}[ht]
    % \vspace{-5pt}
    \begin{minipage}[b]{\textwidth}
        % \begin{center}

        \subfigure[MHC aligned pangenome]{
            \includegraphics[width=.6\textwidth]{fig/mhc-align.eps}
            \label{fig:1a}
        }
        \vspace{-11pt}
        % \newline
        \subfigure[MHC consensus pangenome]{
            \includegraphics[width=.4\textwidth]{fig/mhc-pangenome.eps}
            \label{fig:1b}
        }
        \caption{Pangenome visualisation of the Major Histocompatibility Complex (MHC) \textit{locus} by \odgi.
            (a) projection of full MHC \textit{locus} of haploid phased human genome
          assemblies, plus the chm13 cell line and GRCh38 reference genome as supplied by the
            HPRC and created by \cmd{odgi build+sort+viz}. The coloured
            bars represent the linearised paths --- representing contigs as a zoomed out multi-sequence
            alignment. The black lines at the bottom represent the graph topology.
            (b) consensus graph representation by \cmd{odgi build} of the same assemblies showing variations larger than $100$ base pairs.
            Loops or `bubbles' display contigs uniquely diverging from the consensus, caused by variation in paralogs/repeats.
            The gene labels are super imposed by \cmd{odgi position} from a GRCh38 BED file and visualised by the Bandage tool~\citep{Wick:2015}.
        }
        \label{fig:1}
        % \end{center}
    \end{minipage}
\end{figure*}

### 2021-06-02 manuscript/main.tex
edits

\newcommand{\FIXME}{\textbf{!!FIXME!!}}
\newcommand{\cmd}[1]{{\scriptsize\textrm{#1}}}
\newcommand{\cmdbf}[1]{{\textbf{#1}}}
\newcommand{\topic}[1]{{\cmdbf{#1}}:}
\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    \title[ODGI: scalable tools for pangenome graphs]{ODGI: scalable tools for pangenome graphs}
    \author[Guarracino, Heumos \textit{et~al}.]{
        Andrea Guarracino\,$^{\text{\sfb 1,}\dagger,*}$,
        Simon Heumos\,$^{\text{\sfb 2,}\dagger}$,
    % XXXXX,
        Pjotr~Prins\,$^{\text{\sfb 3}}$ and
        Erik~Garrison\,$^{\text{\sfb 3,}*}$
    }
    \address{$^{\text{\sf 1}}$Biology, University of Tor Vergata, Rome, 00133, Italy \\
        $^{\text{\sf 2}}$QBiC, University of Tübingen, Tübingen, XXXXXXXXXX, Germany \\
        $^{\text{\sf 3}}$Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
    }

    \corresp{$^\ast$To whom correspondence should be addressed.}
    \corresp{$^\dagger$The authors wish it to be known that the first two authors should be regarded as joint First Authors.}

@@

    \abstract{ \textbf{Motivation:} Pangenomes address the
      shortcomings of mainline genomics workflows, in particular
      concerning reference genome bias and identification of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that makes use of an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes at the same time while maintaining fast
        execution runtime and low memory overheads.  \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci},
        exploratory analysis, sorting, removing artifacts, and graph
        manipulation and visualisation.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and user manual can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu\\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME}

    \maketitle

@@

    Thanks to advances in sequencing technologies, new genome
    assemblies are produced at a high rate, including the human
    pangenome reference consortium (HPRC) and telomere-to-telemere
    efforts, e.g.~\citep{Miga:2020}.  A pangenome models the full set
    of genomic elements in a given species or clade. Pangenomics
    therefore contrasts with reference genome based approaches which
    relate sequences to a particular consensus model of the
    genome~\citep{Eizenga:2020}. A pangenome can elegantly be
    represented by a graph data structure incorporating sequences as
    nodes and their connections as edges. These nodes are shared for
    homologs, paralogs and orthologs.  A `bidirected' graph represents
    both strands of DNA and so-called paths represent contigs that run
    through the nodes. A path can represent an individual chromosome
    or part thereof. Such a graph-based approach is particularly
    relevant for highly repetitive regions. Segmental duplications,
    and acrocentric chromosomes lead to graphs which present very high
    node depths which greatly increases processing complexity.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
    large data (see Table~\ref{tab:02}). With \odgi\ we
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
    graph structure in RAM that can be dynamically updated using
    multiple CPUs.

    Note that \odgi\ is focused on `first-order' graph
    operations.  Other tools are required to construct graphs from raw
    sequences, map reads, or call variants.  \odgi\ provides
    operations on an existing graph, such as subset, subdivide, break,
    combine, normalize and order its components, nodes, and
    paths. Most of the tools are designed to be applied together,
    piping the output of one tool into the next.

    \vspace{-0.1in}

@@

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, such as
    validating that paths respect the graph's topology.

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome and is
    capable of handling full length human chromosomes
    (Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
    simplifies the graph by applying several algorithms, including the
    novel `Path-Guided Stochastic Gradient Descent' algorithm.
    \cmd{odgi layout} generates a file with the X-Y coordinates for
    all nodes and \cmd{odgi draw} draws a layout of the graph
    (Fig.~\ref{fig:1}b).

@@

    \topic{odgi position} pangenome graphs are flexible when it comes
    to coordinate systems. \cmd{odgi position} can use the coordinate
    system from a contained reference genome --- a dynamic liftover
    --- to display coordiates and other localised features, as is shown in
    Fig.~\ref{fig:1b}.

        %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[h!]
        \caption{ Example of performance improvement between
            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
            commands with output file sizes. The centromere region was
            extracted from CHR 6 of the HPRC year one assembly, i.e.,
            88~haploid, phased human genome assemblies from
            44~individuals plus the chm13 cell line and GRCh38
            reference genomes stored in 4.3GB~GFA format\citep{GFA}.
            %
            First GFS was converted from GFA to \odgi\ and
            \vg\ formats, respectively.  \odgi's path implementation
            can process paths in parallel, outperforming \vg. For the
            same reason \cmd{odgi extract} is $20\times$ faster than
            \cmd{vg chunk} extracting a portion from the full
            graph. \FIXME: vg never finished.
            %
            All timings were performed on a server class machine
            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME}
        \begin{tabular}{{@{}lllll@{}}}
          \toprule
          Operation & Command & Runtime & Memory  & File size     \\
                    &      & (mm:ss) & (GB) & (GB)           \\
          \midrule
          Covert GFA to native format                         &       &       &                   \\
          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
          Extract subgraph                         &       &       &                   \\
          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
          \botrule
        \end{tabular}
        \label{tab:02}

@@

    Pangenome graphs stand to become a ubiquitous tool in
    genomics\citep{Eizenga:2020}.  With \odgi\ we implemented a
    state-of-the-art tool suite, that can transform, analyse and
    visualise pangenome graphs at scale.  \odgi\ also exposes an API
    that is used by other tools for pangenome graph building and
    normalization. Future work will add support for RNA and protein
    sequences and expand on metadata capabilities of large pangenome
    graphs.

@@

>>>>

\newcommand{\FIXME}{\textbf{!!FIXME!!}}
\newcommand{\cmd}[1]{\textbf{\textrm{#1}}}
\newcommand{\cmdbf}[1]{{\textbf{#1}}}
\newcommand{\topic}[1]{{\cmdbf{#1}}:}
\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    \title[ODGI: scalable tools for pangenome graphs]{ODGI: scalable tools for pangenome graphs}
    \author[Guarracino, Heumos \textit{et~al}.]{
        Andrea Guarracino\,$^{\text{\sfb 1,}*}$,
        Simon Heumos\,$^{\text{\sfb 2,}*}$,
    % XXXXX,
        Pjotr~Prins\,$^{\text{\sfb 3}}$ and
        Erik~Garrison\,$^{\text{\sfb 3,}\dagger}$
        \footnote{The authors wish it to be known that the first two authors should be regarded as joint first authors.}
    }
    \address{$^{\text{\sf 1}}$Biology, University of Tor Vergata, Rome, 00133, Italy \\
        $^{\text{\sf 2}}$QBiC, University of Tübingen, Tübingen, XXXXXXXXXX, Germany \\
        $^{\text{\sf 3}}$Dept. Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
    }

    \corresp{$^\ast$To whom correspondence should be addressed.}

@@

    \abstract{ \textbf{Motivation:} Pangenomes graphs address some of the
      shortcomings of mainline genomics, in particular
      with regard to reference genome bias and identification of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that implements new algorithms and has an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes in memory with fast
        parallel execution.  \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci},
        exploratory analysis, sorting, removing artifacts, and graph
        manipulation and visualisation. \odgi\ also provides an API that allows
        for writing new tool.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and documentation can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu
        % \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
        }
    \maketitle

@@

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are being produced at a high rate, including
    efforts from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}.  In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for homologs, paralogs and
    orthologs. So-called paths represent contigs that run through the
    nodes. A path can represent an individual chromosome or part
    thereof.  A bi-directed graph can even contain both strands of DNA
    as paths. A graph-based approach is particularly relevant for
    exploring complex variation and highly repetitive regions.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
    large data (see Table~\ref{tab:02}). We
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
    graph structure in computer memory (RAM) that can be dynamically updated using
    multiple CPU cores in parallel.

    Note that \odgi\ is focused on `first-order' graph
    operations.  Other tools are required to construct graphs from raw
    sequences, map reads, or call variants.  \odgi\ provides
    operations on an existing graph, such as subset, subdivide, break,
    combine, normalize and order its components, nodes, and
    paths. Most of the tools are designed to be applied together,
    piping the output from one tool into the next.

    % \vspace{-0.1in}

@@

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
    validate paths respect the graph topology.

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome and is
    capable of handling full length human chromosomes
    (see Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
    simplifies the graph by applying several algorithms, including the
    novel `path-guided stochastic gradient descent' algorithm.
    \cmd{odgi layout} generates a file with the X-Y coordinates for
    all nodes and \cmd{odgi draw} draws a layout of the graph
    (Fig.~\ref{fig:1}b).

@@

    \topic{odgi position} pangenome graphs are flexible when it comes
    to coordinate systems. \cmd{odgi position} can use the coordinate
    system from a contained reference genome --- a dynamic liftover
    --- to display coordinates and other localised features, as is shown in
    Fig.~\ref{fig:1b}. Note that multiple reference genomes can be contained
    in the graph and any contained path can be a reference.

@@

    Pangenome graphs stand to become a ubiquitous tool in
    genomics~\citep{Eizenga:2020}.  With \odgi\ we implemented a
    state-of-the-art tool suite, that can transform, analyse and
    visualise pangenome graphs at scale.  \odgi\ also exposes an API
    that can be used by other tools, e.g., for genotyping, pangenome
    graph building and normalization. Future work will add support for
    RNA and protein sequences and expand on metadata capabilities of
    large pangenome graphs.

@@

        %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[h!]
        \caption{ Example of performance improvement between
            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
            commands with output file sizes. The centromere region was
            extracted from CHR 6 of the HPRC year one assembly, i.e.,
            88~haploid, phased human genome assemblies from
            44~individuals plus the chm13 cell line and GRCh38
            reference genomes stored in 4.3GB~GFA format~\citep{GFA}.
            %
            First GFS was converted from GFA to \odgi\ and
            \vg\ formats, respectively.  \odgi's path implementation
            can process paths in parallel, outperforming \vg. For the
            same reason \cmd{odgi extract} is $20\times$ faster than
            \cmd{vg chunk} extracting a portion from the full
            graph. \FIXME: vg never finished.
            %
            All timings were performed on a server class machine
            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME}
        \begin{tabular}{{@{}lllll@{}}}
          \toprule
          Operation & Command & Runtime & Memory  & File size     \\
                    &      & (mm:ss) & (GB) & (GB)           \\
          \midrule
          Covert GFA to native format                         &       &       &                   \\
          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
          Extract subgraph                         &       &       &                   \\
          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
          \botrule
        \end{tabular}
        \label{tab:02}
